Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated

### 2017-02-23 ms_pre_post.Rmd
Up to 'Fertilisation and post-copulatory inbreeding avoidance' in Methods

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females would also have an opportunity to bias paternity towards unrelated males within their set of chosen mates. Evolution of post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates across multiple matings, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further evolution of pre-copulatory or post-copulatory mate choice strategies. However, if polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all evolve, the long-term dynamics of these three reproductive strategies become difficult to predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, the co-occurence of pre-copulatory and post-copulatory inbreeding avoidance might engender some degree of `degeneracy', defined as a phenomenon by which different elements of a system result in identical outputs [i.e., @Edelman2001]. Consequently, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance functionally redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

@@

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay energetic costs associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models that track the frequencies of alleles underlying multiple reproductive strategies that might influence each other's evolution across generations, and that are enacted among relatives that themselves result from reproductive strategies and inbreeding depression expressed in previous generations, are consequently useful to understand and predict evolutionary outcomes.

@@

We model evolution of polyandry and pre-copulatory and post-copulatory inbreeding avoidance by tracking the dynamics of alleles underlying reproductive strategies enacted by interacting females and males within a small focal population. Individuals are defined as related if they share a common ancestor within the population pedigree (i.e., non-zero kinship). Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of the three reproductive strategy traits (i.e., 30 total diploid loci): tendency for polyandry ($P_{a}$), pre-copulatory inbreeding avoidance ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding avoidance ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each of the three traits whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. In overview, each generation proceeds with females paying costs, expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression. Immigrants arrive in the population and density regulation limits population growth. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating. The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding strategies because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with each trait value.

@@

All males in the population are assumed to be available for any female to choose. We therefore assume that there is no opportunity cost of male mating such that mating with one female would reduce a male's availability to mate with any other female. Individual females mate with their total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models a female's total number of different mates rather than solely her total number of matings.

@@

Each male's value with respect to each focal female is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that a focal female mates with the male. Mating is realised by placing each $q^{m}_{ij}$ value into a vector whose elements sum to unity, and a random uniform number is then generated to sample a single male from this vector. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

>>>>

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females would also have an opportunity to bias paternity towards unrelated males within their set of chosen mates. Evolution of post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates across multiple matings, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further evolution of pre-copulatory or post-copulatory mate choice strategies. However, if polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all evolve, the long-term dynamics of these three reproductive strategies become difficult to predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, the co-occurence of pre-copulatory and post-copulatory inbreeding avoidance might engender some degree of 'degeneracy', defined as a phenomenon by which different elements of a system result in identical outputs [i.e., @Edelman2001]. Consequently, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance functionally redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

@@

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay initial energetic costs associated with developing the physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models that track the frequencies of alleles underlying multiple reproductive strategies that might influence each other's evolution across generations, and that are enacted among relatives that themselves result from reproductive strategies and inbreeding depression expressed in previous generations, are consequently useful to understand and predict evolutionary outcomes.

@@

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance, by tracking the dynamics of alleles underlying reproductive strategies enacted by interacting females and males within a small focal population. We thereby tracked evolutionary dynamics given internally-consistent patterns of relatedness caused by non-random mating while also allowing mutation, gene flow, drift, and selection to potentially affect evolution. Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of the three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$), pre-copulatory inbreeding avoidance ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding avoidance ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each of the three traits whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. In overview, each generation proceeds with females paying costs, expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual strategies of pre-copulatory and post-copulatory inbreeding avoidance to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

<!--- Avoided 'used an IBM' to prevent the redundancy of 'used an IBM to model'--->

<!--- I have added a sentence like that suggested, hitting the key evolutionary processes (as all pop-geneticists will recognise these as fundamental, emphasising that we're accounting for all the possible ways evolution can occur). It's tempting to remove 'gene-flow', which we only really use to prevent kinships from getting too high, but without it it sounds as if we're leaving something out. I don't see why this deals with the concern in the intro though -- the sentence just says what we did; it doesn't really explain why we *needed* to do what we did to address the knowledge gap. --->

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding strategies because both negative and positive $M_{p}$ and $F_{p}$ values affect the degree of inbreeding. In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with each trait value.

<!--- One issue with the 'see Discussion' though is that it's not clear what we're telling readers to see the discussion *about*. Were I to read this as reviewer, I would be confused, ''see Discussion? What for? Because you made an assumption in a model, the biological context of which was already explained in the Introduction? (2nd to last paragraph)''. I've tried to provide a bit of context in parentheses, but it makes me nervous catering to any (mistaken) readers who might be confused about what assumptions are used for in models. I'd rather deal with confused reviewers than write a paper that perpetuates some (increasingly frustrating) misunderstandings of modelling and theory.  --->

@@

All males in the population are assumed to be available for any female to choose. We therefore assume that there is no opportunity cost of male mating such that mating with one female does not reduce a male's availability to mate with any other female. Individual females mate with their total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models a female's total number of different mates rather than solely her total number of matings.

@@

Each male's value with respect to each focal female is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that a focal female mates with the male. Mating is realised by placing each $q^{m}_{ij}$ value into a vector whose elements sum to unity. A new vector of identical size is then created where each element equals the summation of all previous elements in the original vector of $q^{m}_{ij}$ values. Because each element in the vector corresponds to a unique male, a random uniform number between zero and one can identify a female's realised mate. This realisation is enacted by identifying the element in the vector that is closest to, but not less than, the random uniform number. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

<!--- Jane, we might want to step back a bit on this one and figure out what exactly it is that is confusing and needs clarifying, as I'm not sure I see what the issue is anymore or, rather, why we're going into so much detail about a very routine procedure in an IBM (or any bit of code, really). This kind of sampling procedure was used in both of our previous IBM papers and I don't recall it being an issue. I have explained the procedure as carefully as I can, but it's getting to the point now where it's essentially turning the specifics of the C programming language into English (i.e., the methods here are now language dependent -- could be done differently in R or Perl or whatever, essentially just to say that we're sampling an element randomly from a weighted vector (I'm still not sure why we can't go back to just saying it that way; actually, I don't see why we can't just stop after ''the value of q^m_ij then defines the probability that a focal female mates with the male'' -- why does this need further explanation in the text?)). --->
